An Analysis of Neighborhood Attributes in Chicago using Eigenvector Centrality

### 2018-04-28 paper/paper.tex
progress

\usepackage{amsthm}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

@@

\date{\today}
\begin{document}
\begin{titlepage}
\maketitle
\thispagestyle{empty}

@@

Network centrality measures have been studied as a way to study how networks are ordered, evolve, and interact.  Several measures of centrality have been studied in order to address a central question \textit{Who is the most important actor in a network}?  While there is no agreement on one measure of centrality, many have been proposed to to address how different networks are structured and what the researcher means by a node's \textit{importance}.  For example, "degree centrality" is a classic measure of centrality and counts how many connections a node has.  A node is more important if it has many neighbors and less important if it has fewer neighbors.  "Closeness centrality" measures the degree to which an individual is near all other individuals in the network and is the inverse of the sum of the shortest distances between each node and every other node in the network.  It has been used to study how long it would take to spread information from a point to all other nodes sequentially.  "Betweenness centrality" quantifies the number of times a node acts as a bridge along the shortest path between two nodes and has been used to quantify the control of information in a social network. \\

In this paper I will focus on "Eigenvector centrality" which is a measure of influence of a node in a network and corresponds to the first eigenvector in the connectivity matrix of a network.  The first researcher to apply the mathematics of eigenvectors to geography was P.R. Gould.  In his paper, On the Geographical Interpretation of Eigenvalues, his goal was less motivated by a specific hypothesis but more of a curiosity to determine if this mathematical structure could uncover a pattern and order in very complex situations.\cite{gould1967geographical}  His hope, which many computational social scientists share, was that underlying complex phenomena might be a mathematical idea that could provide a meaningful geographic interpretation.  To explore this idea, he mapped the road network of Uganda and created a connectivity matrix of this network on a binary scale, 1 indicating if two cities were connected and 0 if they were not.  He calculated the first four eigenvectors of these matrices in 1921 and 1935 and compared the results between the two years.  He found the first eigenvalue, which is centered around the city of Kampala, was by far the most connected town owning to the number of direct linkages and its central location.  He noticed the city with the next highest value, Entebbe, was also very connected.  He then examines a new connectivity matrix of the cities in 1935 and describes how several structural characteristics have been strengthened as new cities are added to the network and routes between other cities were built.  He notes that the successive eigenvectors and eigenvalues are a “pull out” of small regional networks within the trading structure of the region.  Gould makes an initial attempt, though vague, to describe the meaning of his eigenvector derivation.  He explains that vectors representing well-connected towns will not only lie in the middle of a large number of dimensions but will tend, in turn, to lie close to the principal eigenvalue.  On the other hand, towns that are moderately well connected will not lie in the middle of so many dimensions as the well-connected towns and will tend to form small structural clusters on their own.  This interpretation has been named the “Gould Index of Accessibility”.\\

In a social network represented by an adjacency matrix, Bonacich suggested that eigenvector centrality makes a good centrality measure\cite{bonacich2007some}.  Unlike degree centrality, which weighs every contact equally, eigenvector centrality takes into account the centrality of the neighboring node, so if a node is connected to a more important node it's eigenvector centrality will be higher than if it connected to a less important node.  Because of this, eigenvector centrality takes into account the entire pattern in the network.  Eigenvector centrality also has other benefits such as it can be used for signed graphs, adjacency graphs, or value based graphs.  This measure can also be applied to graphs with negatively connections though I will not explore those in this paper. \\

The importance of centrality has been studied to explain a lot of phenomena in the social sciences and some results have given unique new insights.  For example, Cook et al. have shown that power does not equal centrality in exchange networks.  In negotiations, being in a central position (connected to the most number of people) does not mean an actor is the most powerful in bargaining situations.\cite{cook1983distribution}  Thus, typical centrality measures applied to this type of networks failed to predict power distributions in this type of network and thus display how for some measures their generality is limited.  Through theoretical and simulated results, they were able to show that in bargaining situations, it is advantageous to be connected to those who have few options.  One is more powerful if they are connected to those who are powerless because being connected to other powerful actors reduces ones bargaining power.  In situations where degree centrality fails, other social scientists have proposed other measures.  Two different principals of connection in social networks suggests that current measures of centrality might predict power in one type of network but not another and it offers a step towards a fusion of power-dependence theory and structural centrality in a way which might be general across networks of mboth tyupes\cite{bonacich1987power}

Other researchers have since attempted to derived and explain this phenomena across a number of fields and applications.  For example, Tinkler described these eigenvalues in the context of a rumor spreading through a social network.  He described a social network where people were connected through social ties (1 indicating two people know each other and 0 indicating they do not) and a rumor starting at some vertex $i$ at time 0.  As time progresses, the rumor will be spread throughout the network according to the connections between people in the social network.  If someone knows the rumor they tell it as many times as they heard it to all the people they are in direct contact with.  In his example, a person can also start an "anti-rumor" which is denoted by a negative value which can also propagate through the network.  As this process repeats after a large number of periods, the distribution of the rumor will also be given by the principal eigenvector.   In other words, after time progresses the eigenvector is the chance that the rumor has spread to a specific node in the network.\cite{tinkler1972physical} \\

@@

One last interesting application of eigenvector I will describe was implemented by Sergey Brin and Lawrence Page in creating a web search engine which served as the basis of the first version of Google’s search engine.  They created a database with the hyperlink network of over 24 million pages and a PageRank algorithm to order results of a query.  Pages are arranged in a network based on their hyperlinks to other pages.  PageRank does not count all links equally though, but normalized the weighting by the number of links on the page.  The PageRank of a webpage was calculated using an iterative algorithm that corresponded to the principal eigenvector of the normalized link matrix of the web.  They give a few intuitive justifications why this ranking works.  One was imagining a “random surfer” who is given a web page at random and keeps clicking links.  The PageRank of a page is the probability the random surfer will land on a page.  Another interpretation was that a page will have a high PageRank if there are many pages that point to it, or if there are some pages that point to it and also have a high PageRank so it was a way to combine a ranking that measured reputation and ubiquity.\cite{page1999pagerank}\\

While there have been several explanations of eigenvector centrality in networks across many fields, it is often difficult to interpret the meaning of exactly what an eigenvector corresponds to in the real world.  For example, many of the examples explain how an eigenvector as a measurement of a flow or ranking in equilibrium.  However, what does a future equilibrium state say about the current state.  How long will it take to reach the equilibrium state?  What would happen if there is a shock to the network, how would this equilibrium change?  \\

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=1.0\textwidth]{arc-1}
    \caption{This map shows all the connections between where people live and work in Chicago.}
    \label{fig:arc-1}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=1.0\textwidth]{arc-2}
    \caption{This map shows the connections between a people living Hyde Park who either commute to the University of Chicago or the The Loop}
    \label{fig:arc-2}
\end{figure} \\

@@

Real estate listing data is provided by Zillow, an online real estate database company that tracks detailed listing data of home sales throughout the US starting in 1994.  They publish data of home sales for the top 50 markets on the neighborhood level.  In my paper I use the following statistics they publish:
\begin{enumerate}
\item Median List Price Per Square Foot (\$)
\item Monthly Home Sales (Number, Seasonably Adjusted)
\item Age of Inventory (Days)
\item Median Rent List Price (\$), 1-Bedroom
\end{enumerate}
Since I am comparing real estate prices to neighborhood rankings, I make the following adjustments to be able to compare the data.  First I normalize each neighborhood to it's initial value to get an index of the statistic over time.  Next I divide by the median value for all neighborhoods in Chicago for the given time period.  The result is a statistic that shows how the neighborhood compares to other neighborhoods in Chicago over time.  If the value is less than 1, then it is below average whereas if it is greater than 1 it is above average. \\
For a complete overview of how data was collected, cleaned, visualized, and calculated, I provide the scripts and programs I used on github at \url{https://github.com/b-nroths/chi-data}.

\section{Methods}
In order to find the employment centers of a city I will use a setup based on eigenvector centrality that has been used in studying networks such as trade routes, social networks, and webpages.  The goal of the method is to take a network of linked entities and to output the most influential nodes in the network graph.  The definition of influential varies depending on the context of how the problem is set up and what the nodes and edges represent.  For example, in a social network, the most important node is the person who is connected to the most important people or in the “center”.  In a network of cities that trade with each other, the edges might represent the distance between cities.  The edges can either represent a connection, in which case the matrix is called an adjacency matrix and is filled with 1 and 0’s depending on if two nodes are connected to each other or the matrix could represent the strength of connection between two nodes, for example in the trading routes example it could be the distance between two cities.  In this case it is common to normalize the distances to the sum of each column is equal to one. \\

Before I demonstrate how the network matrix is built in the context of this paper, it is important to understand why we are guaranteed a positive eigenvalue from the following theorem proven by Oskar Perron and Georg Forbenius.

@@

From this information we can create a commuting matrix which normalizes the flow between regions of the city and transforms the “live” matrix (2) into the “work” matrix (3).  This transforms the above matrices into the following equation.

@@

\begin{center}C $l$ = $w$\end{center}
\begin{equation} 
  \begin{bmatrix}
    0.5 & 0.1\\
    0.5 & 0.9
  \end{bmatrix}
  \begin{bmatrix}
    10\\
    100
  \end{bmatrix}
  = 
  \begin{bmatrix}
    15\\
    95
  \end{bmatrix}
\end{equation}

Lastly, I will create a network from the dataset which represents a theoretical economy within the city.  Instead of people, I consider the flow of money between regions modeled by the salary of workers that commute between regions.  The first model will be the total flow of money between regions represented by the sum of the salaries of all the workers who commute between regions.  For example, if 10 people commute from Hyde Park to Downtown and they each make an average of \$2,500 per month I will consider the money flow between Hyde Park and downtown to be \$25,000.  In my dataset, salaries are broken down into three ranges less than \$1,250, \$1,250-\$3,333 and over \$3,333.  To simplify the problem, I will represent the buckets as \$1,250, \$2,500, and \$5,000.  Consider the same commuting flows as above but now with income added according to the following breakout.\\

@@

These results will demonstrate a few things.  First, note that the principle eigenvalue is 1.0 as expected from the theorem above.  The vector that corresponds to the principle eigenvalue is the Gould Index of Accessibility in the network and represents the relative strength of each node in the principle network.  In the three examples above you can see that the Downtown area dominates this vector.  The second eigenvalue represents a "pull out" of the principle network.  It is important to remember that the second eigenvector will be orthogonal to the principle eigenvector and represents a completely different sub-network.  In the system in equation 5 and 7 the second eigenvalue is lower which indicates the dominance of the first eigenvalue compared to the network examined in the 6 equation.  These results have a similar interpretation to the previous example where the principle eigenvector show the most dominant network of living regions.  The eigenvector that corresponds to the principle eigenvalue are the relative ranking of regions in this setting.  \\

@@

Here not only is the primary eigenvector more equal across neighborhoods with values of $0.8087$ and $0.5882$, the value of the second eigenvalue is also very small compared to the first $\lambda_2$ = 0.05.  This means that the majority of the network is explained by it's primary eigenvalue. \\
 Next I will use this same set up to analyze the neighborhoods of Chicago using the LEHD dataset.

>>>>

\usepackage{amsthm}
\usepackage{blindtext}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

@@

\date{\today}
\begin{document}
\begin{titlepage}
\maketitle
\begin{abstract}
For over a century, fingerprints have been an undisputed personal identifier.  Recent court rulings have sparked interest in verifying unique ...
\end{abstract}

\thispagestyle{empty}

@@

Network centrality measures have been studied as a way to analyze how actors are ordered, evolve, and interact.  Several measures of centrality have been studied in order to address a central question \textit{Who is the most important actor in a network}?  While there is no agreement on one measure of centrality, many researchers suggested centrality measurements based on how a specific network is structured, actor importance, actor relationships, and what the researchers mean by ``importance''.  For example, ``degree centrality'' is a classic measure of centrality that counts how many connections an actor has.  A node is more important if it has many neighbors and less important if it has fewer neighbors.  ``Closeness centrality'' measures the degree to which an individual is near all other individuals in the network and is the inverse of the sum of the shortest distances between each node and every other node in the network.  It has been used to study how long it would take to spread information from one actor to all others sequentially.  ``Betweenness centrality'' quantifies the number of times a node acts as a bridge along the shortest path between two nodes and has been used to quantify the control of information in a social network. \\

In this paper I will focus on ``Eigenvector centrality'' which is a measure of influence of a node in a network and corresponds to the first eigenvector in the connectivity matrix of a network.  The first researcher to apply the mathematics of eigenvectors to geography was P.R. Gould.  In his paper, On the Geographical Interpretation of Eigenvalues, his goal was less motivated by a specific hypothesis but more of a curiosity to determine if this mathematical structure could uncover a pattern and order in very complex situations.\cite{gould1967geographical}  His hope, which many computational social scientists share, was that underlying complex phenomena might be a mathematical idea that could provide a meaningful geographic interpretation.  To explore this idea, he mapped the road network of Uganda and created a connectivity matrix of this network on a binary scale, 1 indicating if two cities were connected and 0 if they were not.  He calculated the first four eigenvectors of these matrices in 1921 and 1935 and compared the results between the two years.  He found the first eigenvalue, which is centered around the city of Kampala, was by far the most connected town owning to the number of direct linkages and its central location while the city with the next highest value, Entebbe, was also very connected.  He then examined a new connectivity matrix of the cities in 1935 and described how several cities have become more important as new roads and cities have been built.  He notes that the successive eigenvectors and eigenvalues are a “pull out” of small regional networks within the trading structure of the region.  Gould makes an initial attempt, though vague, to describe the meaning of his eigenvector derivation.  He explains that vectors representing well-connected towns will not only lie in the middle of a large number of dimensions but will tend, in turn, to lie close to the principal eigenvalue.  On the other hand, towns that are moderately well connected will not lie in the middle of so many dimensions as the well-connected towns and will tend to form small structural clusters on their own.  This interpretation has been named the “Gould Index of Accessibility”.\\

Eigenvector centrality has been studied to explain a lot of phenomena in the social sciences and some results have given new insights when previous theories and measures of centrality failed.  For example, Cook et al. argued that typical centrality measures like degree centrality have failed to predict power distributions in exchange networks (networks where actors bargain or trade which each other).\cite{cook1983distribution}  Through theoretical and simulated results they showed that in negotiations it is advantageous to be connected to those who have few options and being in a central position does not make an actor more powerful.  This is because if an actor is connected to someone who is powerless they will have more negotiating power because if the powerless actor was connected to other powerful actors, their bargaining power would increase.  Thus, they suggest that a more general conception of point centrality needs to be developed that takes into account power dependency as well as closeness.  In situations where degree centrality fails, social scientists have proposed other measures that fuse power-dependency and closeness.  Bonacich suggested that eigenvector centrality makes a good centrality measure in these networks because it takes into account the number of connections an actor has as well as the centrality of the actor one is negotiating with.\cite{bonacich1987power}  Thus, if an actor is connected to a more important node it's eigenvector centrality will be higher than if it connected to a less important actor.\footnote{bonacich2007some}  Two different principals of connection in social networks suggests that current measures of centrality might predict power in one type of network but not another and it offers a step towards a  in a way which might be general across networks of both types.  Eigenvector centrality also has benefits over other measures of centrality as it can be used for signed graphs, adjacency graphs, or value based graphs.  For example, networks graphs with negative connections include dating and friendship networks where reciprocation is not necessary or trade where one actor sells a product to another. \\

Other researchers have since attempted to explain eigenvector centrality across a number of fields and applications.  For example, Tinkler described these eigenvalues in the context of a rumor spreading through a social network.  He described a social network where people were connected through social ties (1 indicating two people know each other and 0 indicating they do not) and a rumor starting at some vertex $i$ at time 0.  As time progresses, the rumor will be spread throughout the network according to the connections between people in the social network.  If someone knows the rumor they tell it as many times as they heard it to all the people they are in direct contact with.  In his example, a person can also start an "anti-rumor" which is denoted by a negative value which can also propagate through the network.  As this process repeats after a large number of periods, the distribution of the rumor will also be given by the principal eigenvector.  In other words, after time progresses the eigenvector is the chance that the rumor has spread to a specific node in the network.\cite{tinkler1972physical} \\

@@

Eigenvectors are also commonly used to analyze population flows over time.  For example, imagine a scenario with two cities where a constant proportion of the population moves between cities at every time period.  The key to analyzing how this is in an equilibrium state is through it's eigenvalues and eigenvectors.  This is because the information in the eigenvectors and eigenvalues contain all information needed to study how a system moves from one state to the next.\\

One last interesting application of eigenvector centrality was implemented by Sergey Brin and Lawrence Page in creating a web search engine which served as the basis of the first version of Google’s search engine.  They created a database with the hyperlink network of over 24 million pages and a PageRank algorithm to order results of a query.  Pages are arranged in a network based on their hyperlinks to other pages.  PageRank does not count all links equally though, but normalized the weighting by the number of links on the page.  The PageRank of a webpage was calculated using an iterative algorithm that corresponded to the principal eigenvector of the normalized link matrix of the web.  They give a few intuitive justifications why this ranking works.  One was imagining a “random surfer” who is given a web page at random and keeps clicking links.  The PageRank of a page is the probability the random surfer will land on a page.  Another interpretation was that a page will have a high PageRank if there are many pages that point to it, or if there are some pages that point to it and also have a high PageRank so it was a way to combine a ranking that measured reputation and ubiquity.\cite{page1999pagerank}\\

While there have been several explanations of eigenvector centrality in networks across many fields, it is often difficult to interpret the meaning of exactly what an eigenvector corresponds to in the real world.  For example, many of the examples explain how an eigenvector as a measurement of an exchange or ranking in equilibrium.  However, what does a future equilibrium state say about the current state.  How long will it take to reach the equilibrium state?  What would happen if there is a shock to the network, how would this equilibrium change?  \\

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\textwidth]{arc-1}
    \caption{This map shows all the connections between where people live and work in Chicago.}
    \label{fig:arc-1}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\textwidth]{arc-2}
    \caption{This map shows the connections between a people living Hyde Park who either commute to the University of Chicago or the The Loop}
    \label{fig:arc-2}
\end{figure} \\

@@

Real estate listing data is provided by Zillow, an online real estate database company that tracks detailed listing data of home sales throughout the US starting in 1994.  They publish data of home sales for the top 50 markets on the neighborhood level.  In my paper I use the following statistics:
\begin{enumerate}
\item Median List Price Per Square Foot (\$)
\item Monthly Home Sales (Number, Seasonably Adjusted)
\item Age of Inventory (Days)
\item Median Rent List Price (\$), 1-Bedroom
\end{enumerate}
Since I am comparing real estate prices to neighborhood rankings, I make the following adjustments to be able to compare the data.  First I normalize each neighborhood to it's initial value to get an index of the statistic over time.  Next I divide by the median value for all neighborhoods in Chicago for the given time period.  The result is a statistic that shows how the neighborhood compares to other neighborhoods in Chicago over time.  If the value is less than 1, then it is below average whereas if it is greater than 1 it is above average. \\
For a complete overview of how data was collected, cleaned, visualized, and calculated, I provide the scripts and programs I used on github at \url{https://github.com/b-nroths/chi-data}.

\section{Methods}
In order to find the employment centers of a city I will represent the city's commuting network as a matrix and calculate the important neighborhoods using eigenvector centrality, a method that has been used in studying networks such as trade routes, social networks, and webpages.  The goal of the method is to take the network matrix and to output the most influential actors in the network.  The definition of influential varies depending on the context and will be explore later in this paper.  The matrix representing the connections between actors can be represented a number of way.  In social networks, edges represent a connection, in which case the matrix is called an adjacency matrix and is filled with 1 and 0’s depending on if two nodes are connected to each other.  In a round-robin tournament the matrix could be filled in with 1's and 0's depending on if a team beat the other or by a percentage which would represent a team's win percentage against another team.  In a trading network there are multiple ways to describe the network matrix.  One simple way is to to use an adjacency matrix (as Gould did) of 1's and 0's if an actor trades with another.  Another way would be to represent the size of the actor compared to the country they are trading with or the percent of their total trade an actor makes with a specific partner.  Further it could also be represented as a distance of a trading route between two actors.  In this case it is common to normalize the distances to the sum of each column is equal to one. \\

In this paper I will call the network matrix a ``commuting matrix''.  The actor is a neighborhood and the value in the matrix represents the percent of people who live in one neighborhood who commute to the other.  This will have the benefit of producing an easy to interpret eigenvector and eigenvalue.  Before I demonstrate how the network matrix is built in the context of this paper, it is important to understand why we are guaranteed a positive eigenvalue from the following theorem proven by Oskar Perron and Georg Forbenius.

@@

From this information we can create a commuting matrix which normalizes the flow between regions of the city and transforms the ``live" matrix (2) into the ``work" matrix (3).  This transforms the above matrices into the following equation.

@@

\begin{center}C $l$ = $w$\end{center}
\begin{equation} 
  \begin{bmatrix}
    5/10 & 10/100\\
    5/10 & 90/100
  \end{bmatrix}
  \begin{bmatrix}
    10\\
    100
  \end{bmatrix}
  = 
  \begin{bmatrix}
    15\\
    95
  \end{bmatrix}
\end{equation}

The last connectivity matrix I will study from this dataset represents a theoretical economy within the city.  Instead of people, I consider the flow of money between regions modeled by the salary of workers that commute between regions.  The first model will be the total flow of money between regions represented by the sum of the salaries of all the workers who commute between regions.  For example, if 10 people commute from Hyde Park to Downtown and they each make an average of \$2,500 per month I will consider the money flow between Hyde Park and downtown to be \$25,000.  In my dataset, salaries are broken down into three ranges less than \$1,250, \$1,250-\$3,333 and over \$3,333.  To simplify the problem, I will represent the buckets as \$1,250, \$2,500, and \$5,000.  Consider the same commuting flows as above but now with income added according to the following breakout.\\

@@

These results will demonstrate a few things.  First, note that the principle eigenvalue is 1.0 as expected from the theorem above.  The vector that corresponds to the principle eigenvalue is the Gould Index of Accessibility in the network and represents the relative strength of each node in the principle network.  In the three examples above you can see that the Downtown area dominates this vector.  The second eigenvalue represents a ``pull out" of the principle network.  It is important to remember that the second eigenvector will be orthogonal to the principle eigenvector and represents a completely different sub-network or clique.  In the system in equation 5 and 7 the second eigenvalue is lower which indicates the dominance of the first eigenvalue compared to the network examined in the 6 equation.  These results have a similar interpretation to the previous example where the principle eigenvector show the most dominant network of living regions.  The eigenvector that corresponds to the principle eigenvalue are the relative ranking of neighborhoods in this setting.  \\

@@

Here not only is the primary eigenvector more equal across neighborhoods with values of $0.8087$ and $0.5882$, the value of the second eigenvalue is also very small compared to the first $\lambda_2$ = 0.05.  This means that the majority of the network is explained by it's primary eigenvalue. \\

Next I will use this same set up to analyze the neighborhoods of Chicago using the LEHD dataset.
